Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms

### 2011-09-14 docs/thesis/paper.tex
Updated paper with pseudocode and new figures

\documentclass{article}
\usepackage{graphicx}
\begin{document}
\title{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}
\author{Alvaro Prieto}
\maketitle

@@

There are several routing protocols designed specifically for WBANs. Braem et al. developed the CICADA protocol, which consists of a spanning tree architecture with a time-division scheme for transmission scheduling~\cite{protocol:CICADA}. The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes. Other protocols focus on the reliability of the network~\cite{routing:storeandforward}. Dynamic networks are a problem with WBANs. If sensors are placed in the extremities, the network will change as the body changes positions. Quwaider et al. developed a protocol tolerant to network changes~\cite{routing:storeandforward}. They propose a store-and-forward method that maximizes the likelihood of a packet reaching its destination. Each packet is stored by multiple devices and retransmitted, which consumes more power. One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources. While this method increases the network lifetime$\ll$define term$\gg$, it requires more hardware. Finally, Nabi et al. propose a similar store-and-forward method, but integrates Transmit Power Adaptation (TPA)~\cite{relay:transmitpoweradaptation}. Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/body.pdf}
\end{center}
\caption{Experimental Setup}
\label{fig:body}
\end{figure}

@@

Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI. The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$. The channel attenuation for the selected link,$\alpha$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power and $P_tx$ is the transmitted power. The actual link cost is computed by calculating the energy used by a device \emph{if} that link is selected and subtracting the average energy used across all devices. Equation~\ref{eqn:link_cost} is normalized by the current polling round $i$.
\begin{equation}
\label{eqn:alpha}
\alpha = \frac{RSSI}{P_{tx}}
\end{equation}
\begin{equation}
\label{eqn:accumulated_energy}
E^j_{(i)} = E^j_{(i-1)} + \frac{RSSI^T}{\alpha}
\end{equation}
\begin{equation}
\label{eqn:link_cost}
C^j_{l,k} = \left[ \left(\frac{ E^l_{(i-1)} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{(i-1)}}{i-1} \right]
\end{equation}

\begin{equation}
\label{eqn:link_cost}
C^j_{l,k} = \frac{RSSI^T}{\alpha_{l,k}} \times \left( \frac{1+\left(\frac{E^j_i}{E^{min}_{i}} \right)^c}{2} \right)
\end{equation}

@@

This algorithm works under several assumptions. The first is that this is a centralized network with one central Access Point (AP) with virtually unlimited energy and  computing power. The second is that each wireless sensor, or End Device(ED), can act as a relay for messages from other nodes. The Third and last requirement is that the link quality,or channel attenuation, between all EDs and AP is known. All routing calculations take place in the AP.

$\ll$describe network (centralized network with powerful AP and multiple sensors able to act as relays, power control), figure 1 (like in paper fig.1 + symbols for channel attenuation, cost, sensor ID,…) and describe symbols in figure.$\gg$

$\ll$Define accumulated energy (normalized power), cost function and all relevant symbols$\gg$

@@

\section{Performance Analysis}
$\ll$”in what follows we …. (state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$

@@

@@

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/walk2-c100.pdf}
\end{center}
\caption{Walking (C=100)}
\label{fig:walk2-c100}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/walk2-c0.pdf}
\end{center}
\caption{Walking (C=0)}
\label{fig:walk2-c0}
\end{figure}

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\begin{document}
\title{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}
\author{Alvaro Prieto}
\maketitle

@@

There are several routing protocols designed specifically for WBANs. Braem et al. developed the CICADA protocol, which consists of a spanning tree architecture with a time-division scheme for transmission scheduling~\cite{protocol:CICADA}. The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes. Other protocols focus on the reliability of the network~\cite{routing:storeandforward}. Dynamic networks are a problem with WBANs. If sensors are placed in the extremities, the network will change as the body changes positions. Quwaider et al. developed a protocol tolerant to network changes~\cite{routing:storeandforward}. They propose a store-and-forward method that maximizes the likelihood of a packet reaching its destination. Each packet is stored by multiple devices and retransmitted, which consumes more power. One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources. While this method increases the network lifetime, it requires more hardware. Finally, Nabi et al. propose a similar store-and-forward method, but integrates Transmit Power Adaptation (TPA)~\cite{relay:transmitpoweradaptation}. Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.

For the purpose of this paper, \textbf{network lifetime} is defined as the time it takes a single node to deplete its energy source.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/body.pdf}
\end{center}
\caption{Experimental Setup (Not all links shown)}
\label{fig:body}
\end{figure}

@@

\subsection{Device Energy}
Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha_{j,k}$ is the channel attenuation for the selected link, and $RSSI^T$ is the target RSSI. The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$. The channel attenuation for the selected link,$\alpha_{j,k}$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power measured by device $k$ and $P_{tx}$ is the transmitted power used by device $j$.

\subsection{Link Cost}
The actual link cost is computed by calculating the energy used by the device, \emph{if} that link is selected, and multiplying by a cost factor(Equation~\ref{eqn:link_cost}). The cost factor calculation takes the energy used by the destination device, $E^k_i$, and divides it by the current energy minimum, $E^{min}_{i}$, which is the energy used by the device with the minimum energy used by round $i$. This ratio is then raised to the power $M$, which determines how strong the effect will be $\ll$Need a better way of saying this..$\gg$. If a device's energy is much greater than the current minimum, it will be avoided as a relay.
The $M$ factor in Equation~\ref{eqn:link_cost} can be any number in $0 < M < \infty$. The cost factor is normalized so when $M=0$, it goes back to the traditional cost of required power to meet the link.

@@

\begin{equation}
\label{eqn:alpha}
\alpha_{j,k} = \frac{RSSI}{P_{tx}}
\end{equation}
\begin{equation}
\label{eqn:accumulated_energy}
E^j_{(i)} = E^j_{(i-1)} + \frac{RSSI^T}{\alpha}
\end{equation}
\begin{equation}
\label{eqn:link_cost}
C^j_{j,k} = \frac{RSSI^T}{\alpha_{j,k}} \times \left( \frac{1+\left(\frac{E^k_i}{E^{min}_{i}} \right)^M}{2} \right)
\end{equation}

\subsection{Assumptions}
This algorithm works under several assumptions. The first is that this is a centralized network with one central Access Point (AP) with virtually unlimited energy and  computing power. The second is that each wireless sensor, or End Device(ED), can act as a relay for messages from other nodes. The Third is that the link quality,or channel attenuation, between all EDs and AP is known. Finally, all EDs begin with an equal energy used, that is, they all have the same energy stored. All routing calculations take place in the AP. Figure~\ref{fig:body} shows a sample network with EDs labeled 1 through 8, and one AP.

\subsection{Performance Metric}
The goal of the new cost function is to increase network lifetime.

@@

\section{Performance Analysis}
In what follows, we analyze the performance of the proposed cost function. 
$\ll$”in what follows we …. (state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$

\subsection{Algorithm Implementation}
\begin{algorithm}
\caption{Dijkstra's Algorithm with new cost function}
\label{alg:dijkstra}
\begin{algorithmic}
\FOR{$i = 1$ to Total Nodes} 
	\STATE $node_i.visited \leftarrow 0$
	\STATE $node_i.distance \leftarrow \infty$
\ENDFOR
\STATE $node_{AP}.distance \leftarrow 0$
\STATE $energy_{min} = min(node.energy)$
\WHILE{unvisited nodes available}
	\STATE $node_{source} \leftarrow $ node with smallest distance
	\FOR{$i = 1$ to Total Links}
		\IF{$link_i.source = node_{source}$}
			\STATE $node_{dest} = link_i.destination$
			\IF{$node_{dest}.visited = 0$}
				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
				\STATE $newdistance \leftarrow node_{source}.distance + cost$
				\IF{$distance < node_{dest}.distance$} 
				\STATE$node_{dest}.distance \leftarrow newdistance$
				\ENDIF
			\ENDIF
		\ENDIF
	\ENDFOR
	\STATE $node_{source}.visited \leftarrow 1$
\ENDWHILE
\end{algorithmic}
\end{algorithm}

@@

Two experimental setups were used. For the first, the subject remained relatively static in a sitting position. The second consisted of a more dynamic environment, where the subject walked around a room.

@@

\begin{figure}[htb]
\label{tab:rssi}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
\hline 
\end{tabular} 
\caption{Sample RSSI Table}
\end{figure}

@@

Figure~\ref{fig:sit1plots} shows the resulting accumulated energies while the subject was in a sitting position. The experiment was run with both an $M=0$ as a reference, and $M=100$.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=1.0\textwidth]{figures/sit1plots.pdf}
\end{center}
\caption{Sitting experiment with M=0 and M=100}
\label{fig:sit1plots}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=1.0\textwidth]{figures/walk2plots.pdf}
\end{center}
\caption{Walking experiment with M=0 and M=100}
\label{fig:walk2plots}
\end{figure}
